---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```


# collinear: R Package for Seamless Multicollinearity Management


<!-- Development badges 

[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)
[![Devel-version](https://img.shields.io/badge/devel%20version-1.0.1-blue.svg)](https://github.com/blasbenito/collinear)



<!-- badges: start -->

[![DOI](https://zenodo.org/badge/698046617.svg)](https://zenodo.org/doi/10.5281/zenodo.10039396)
[![CRAN status](https://www.r-pkg.org/badges/version/collinear)](https://cran.r-project.org/package=collinear)
[![CRAN\_Download\_Badge](http://cranlogs.r-pkg.org/badges/grand-total/collinear)](https://CRAN.R-project.org/package=collinear)
[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)
[![lifecycle](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![License](https://img.shields.io/badge/license-MIT-blue.svg)](https://www.gnu.org/licenses/MIT.en.html)

<!-- badges: end -->




## Summary

The R package `collinear` combines four different methods to offer a comprehensive tool for multicollinearity management:

  + **Bivariate correlation for numeric and categorical predictors**: Pearson and Spearman correlation methods for pairs of numeric predictors, and Cramer's V for pairs of categorical predictors.
  + **Variance Inflation Factor analysis (VIF)**: to identify predictors that are linear combinations of other predictors.
  + **Target encoding of categorical predictors**: to convert them to numeric using a numeric variable as response (usually a response variable) and handle them as numerics during the multicollinearity filtering
  + **Flexible feature prioritization**: allows defining the order of importance for the predictors based on specific analytic needs.
  
These methods are integrated in the `collinear()` function, which returns a vector of selected predictors with a controlled multicollinearity.

```{r, eval = FALSE}
selected_variables <- collinear(
  df, #your data frame
  response, #name of your response variable
  predictors, #names of your predictors,
  preference_order, #your predictors in order of interest
  max_cor, #maximum bivariate correlation
  max_vif, #maximum variance inflation factor
  encoding_method, #method to convert categorical predictors into numerics
)
```

The package contains other functions that may be useful during multicollinearity management:
  + `cor_select()`: like `collinear()`, but only using pairwise correlations.
  + `vif_select()`: like `collinear()`, but only using variance inflation factors.
  + `preference_order()`: to compute preference order based on univariate models.
  + `target_encoding_lab()`: to convert categorical predictors into numeric using several methods.
  + `cor_df()`: to generate a data frame with all pairwise correlation scores.
  + `cor_matrix()`: to convert a correlation data frame into matrix, or obtain a correlation matrix.
  + `vif_df()`: to obtain a data frame with all variance inflation factors.
  
## Citation

If you found this package useful during your research work, please cite it as:

*Blas M. Benito (2023). collinear: R Package for Seamless Multicollinearity Management. Version 1.0.1. DOI: 10.5281/zenodo.10039397*

## Install

The package `collinear` can be installed from CRAN.

```{r}
install.packages("collinear")
library(collinear)
```

Three are no big development plans for this package, but still, latest bug fixes and features will surely be in the development version.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "development"
  )
```


## Multicollinearity management with the `collinear` package.

This section shows the basic usage of the package and offers a brief explanation on the methods used within.

```{r, message = FALSE, warning = FALSE}
library(collinear)
library(dplyr)
library(tictoc)
```


### Example data

The package is shipped with a data frame named `vi`, with 30.000 records, and 67 columns with a mixture of types.

```{r}
dplyr::glimpse(vi)
```

The response variables are "vi_mean", "vi_max", "vi_min", and "vi_range", with statistics of a vegetation index named NDVI. The predictors are stored in the character vector `vi_predictors`.

```{r}
vi_predictors
```

### `collinear()`

The `collinear()` function contains the functionality required for a robust multicollinearity management.

#### Input arguments

The function takes these inputs:

  + `df`: a data frame with predictors, and preferably, a response (more about this later).
  + `response`: the name of the response variable, only relevant **and highly recommended** if there are categorical variables within the predictors.
  + `predictors`: names of predictors involved in the multicollinearity analysis.
  + `preference_order`: names of the predictors in the user's order of preference. Does not need to name all predictors in `predictors`!
  + `cor_method`: usually "pearson", but also "spearman" is accepted.
  + `max_cor`: maximum correlation allowed between two predictors.
  + `max_vif`: maximum VIF allowed in a predictor.
  + `encoding_method`: method used to convert categorical variables into numeric. Only relevant when a `response` is provided. By default, each group of the categorical variable is encoded with the mean of the `response` across the group.
  
The code below shows a quick example. Notice that the argument `preference_order` was left as NULL, but will be explained later.

```{r}
selected_predictors <- collinear(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = NULL,
  max_cor = 0.75,
  max_vif = 5,
  encoding_method = "mean"
)

selected_predictors
```

The function has returned a list of predictors that should have a correlation lower than 0.75 with each other, and a VIF lower than 5. Let's see if that's true.

The function `cor_df()` returns a data frame with pairwise correlations, arranged by the absolute value of the correlation. 

```{r}
selected_predictors_cor <- cor_df(
  df = vi,
  response = "vi_mean",
  predictors = selected_predictors
)
head(selected_predictors_cor)
```

The data frame above shows that the maximum correlation between two of the selected predictors is below 0.75, so here `collinear()` worked as expected.

The function `vif_df()` returns a data frame with the VIF scores of all predictors.

```{r}
selected_predictors_vif <- vif_df(
  df = vi,
  response = "vi_mean",
  predictors = selected_predictors
)
selected_predictors_vif
```
The output shows that the maximum VIF is 4.2, so here `collinear()` did its work as expected.

#### Arguments `max_cor` and `max_vif`

The arguments `max_cor` and `max_vif` control the intensity of the multicollinearity filtering.

```{r}
#restrictive setup
selected_predictors_restrictive <- collinear(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  max_cor = 0.5,
  max_vif = 2.5
)

#permissive setup
selected_predictors_permissive <- collinear(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  max_cor = 0.9,
  max_vif = 10
)
```

These are the variables selected under a restrictive setup:

```{r}
selected_predictors_restrictive
```

These are the variables selected under a more permissive setup:

```{r}
selected_predictors_permissive
```

As expected, the restrictive setup resulted in a smaller set of selected predictors. There are no hard rules for `max_cor` and `max_vif`, and their selection will depend on the objective of the analysis and the nature of the predictors.

#### The `response` argument

The response argument is used to encode categorical variables as numeric. When omitted, the `collinear()` function ignores categorical variables. However, the function `cor_select()` can help when there is not a suitable `response` variable in a data frame. This option is discussed at the end of this section.

```{r}
selected_predictors_response <- collinear(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors
)

selected_predictors_no_response <- collinear(
  df = vi,
  predictors = vi_predictors
)
```

When the argument `response` is used, the output may contain categorical predictors (tagged with `<chr>`, from "character" below).

```{r}
dplyr::glimpse(vi[, selected_predictors_response])
```

However, when the argument `response` is ignored, all categorical predictors are ignored.

```{r}
dplyr::glimpse(vi[, selected_predictors_no_response])
```
If there are categorical variables in a data frame, but there is no suitable `response` variable, then the function `cor_select()` can handle the multicollinearity management via pairwise correlations, but at a MUCH higher computational cost, and with different results, as shown below.

```{r}
tictoc::tic()
selected_predictors_response <- cor_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors
)
tictoc::toc()

tictoc::tic()
selected_predictors_no_response <- cor_select(
  df = vi,
  predictors = vi_predictors
)
tictoc::toc()
```

```{r}
selected_predictors_response
```

```{r}
selected_predictors_no_response
```

The variable selection results differ because the numeric representations of the categorical variables are rather different between the two options. When no `response` is provided, the function `cor_select()` compares categorical predictors against numeric ones by encoding each categorical after each numeric, and compares pairs of categoricals using Cramer's V, implemented in the function `cramer_v()`. Additionally, Cramer's V values are not directly comparable with Pearson or Spearman correlation scores, and having them together in the same analysis might induce bias during the variable selection. Not using the `response` argument should always be the last option. 

#### Preference order

The argument `preference_order` gives the user some control on what predictors should be removed first and what predictors should be kept during the multicollinearity filtering. This argument accepts a vector of predictor names in the order of interest, or the result of the function `preference_order()`, which allows to define preference order following a quantitative criteria. 

##### Manual preference order

Let's start with the former option. Below, the argument `preference_order` names several predictors that are of importance for a hypothetical analysis. The `predictors` not in `preference_order` are ranked by the absolute sum of their correlations with other predictors during the pairwise correlation filtering, and by their VIF during the VIF-based filtering.

```{r}
selected_predictors <- cor_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = c(
    "soil_temperature_mean",
    "soil_temperature_max",
    "soil_type"
  )
)

selected_predictors
```

Notice that in the output, two of the variables in `preference_order` are selected ("soil_temperature_mean" and "soil_type"), but one was removed ("soil_temperature_max"). This happens because at some point in the selection, the VIF of "soil_temperature_mean" and "soil_temperature_max" was higher than `max_vif`, and the one with lower preference was removed.

##### Quantitative preference order

The function `preference_order()` requires the `response` argument, and takes a function `f` that returns a value of association between the response and any predictor. This value is then located in the "preference" column of the function's output.

```{r}
preference_rsquared <- preference_order(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  f = f_rsquared,
  workers = 1 #requires package future and future.apply for more workers
)

preference_rsquared
```

The result of `preference_order()` can be plugged right away into the `preference_order` argument of collinear.

```{r}
selected_predictors <- collinear(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = preference_rsquared
)
selected_predictors
```
This variable selection satisfies three conditions at once: maximum correlation between each predictor and the response, maximum pairwise correlation, and maximum VIF.

The `f` argument used by default is the function `f_rsquared()`, that returns the R-squared between the response and any predictor.

```{r}
f_rsquared(
  x = "growing_season_length",
  y = "vi_mean",
  df = vi
)
```

There are two other `f` functions implemented:

  + `f_ga_deviance()`: returns the explained deviance of a univariate GAM model between the response and each predictor, fitted with the function `mgcv::gam()`. Only if the R package `mgcv` is installed in the system.
  + `f_rf_deviance()`: returns the explained deviance of a univariate Random Forest model between the response and each predictor, fitted with the function `ranger::ranger()`. Only if the R package `ranger` is installed in the system.
  
Custom functions created by the user are also accepted as input, as long as they have the `x`, `y`, and `df` arguments, and they return a single numeric value.

These can be run in parallel across predictors by increasing the value of the `workers` argument if the R packages `future` and `future.apply` are installed in the system.

### `cor_select()` and `vif_select()`

The functions `cor_select()` and `vif_select()`, called within `collinear()`, perform the pairwise correlation filtering, and the VIF-based filtering. The main difference between them is that `cor_select()` can handle categorical predictors even when the `response` is omitted, while `vif_select()` ignores them entirely in such case.

```{r}
selected_predictors_cor <- cor_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = preference_rsquared
)

selected_predictors_vif <- vif_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = preference_rsquared
)
```

```{r}
selected_predictors_cor
```

```{r}
selected_predictors_vif
```

### `target_encoding_lab()`

The function `target_encoding_lab()` is used within all other functions in the package to encode categorical variables as numeric. It implements four target encoding methods:

 + "mean" (in `target_encoding_mean()`): replaces each category with the mean of the response across the category. White noise can be added to this option to increase data variability.
 + "rank" (in `target_encoding_rank()`): replaces each category with the rank of the mean of the response across the category.
 + "rnorm" (in `target_encoding_rnorm()`): replaces each value in a category with a number generated by `stats::rnorm()` from a normal distribution with the mean and the standard deviation of the response over the category.
 + "loo" (in `target_encoding_loo()`): replaces each value in a category with the mean of the response across all the other cases within the category. White noise can be added to this option to increase data variability.
 
The method "mean" is used as default throughout all functions in the package, but can be changed via the argument `encoding_method`.
 
Below we use all methods to generate different numeric encodings for the categorical variable "koppen_zone".

```{r}
df <- target_encoding_lab(
  df = vi,
  response = "vi_mean",
  predictors = "koppen_zone",
  encoding_methods = c(
    "mean",
    "rank",
    "rnorm",
    "loo"
  ),
  seed = 1,
  rnorm_sd_multiplier = c(0, 0.01, 0.1),
  white_noise = c(0, 0.01, 0.1),
  verbose = TRUE
)
```

The relationship between these encoded versions of "koppen_zone" and the response are shown below.

```{r, fig.width=9, fig.height=9, eval = FALSE}
#get names of encoded variables
koppen_zone_encoded <- grep(
  pattern = "*__encoded*",
  x = colnames(df),
  value = TRUE
)

#record the user's graphical parameters
user.par <- par(no.readonly = TRUE)

#modify graphical parameters for the plot
par(mfrow = c(4, 3))

#plot target encoding
x <- lapply(
  X = koppen_zone_encoded,
  FUN = function(x) plot(
    x = df[[x]],
    y = df$vi_mean,
    xlab = x,
    ylab = "vi_mean",
    cex = 0.5
    )
)

#reset the user's graphical parameters
par(user.par)
```

The function implementing each method can be used directly as well. The example below shows the "mean" method with the option `replace = FALSE`, which replaces the categorical values with the numeric ones in the output data frame.

```{r}
head(vi[, c("vi_mean", "koppen_zone")], n = 10)
```


```{r}
df <- target_encoding_mean(
  df = vi,
  response = "vi_mean",
  predictor = "koppen_zone",
  replace = TRUE
)

head(df[, c("vi_mean", "koppen_zone")], n = 10)
```

 
