---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
# options(tibble.print_min = 5, tibble.print_max = 5)
```


# collinear: R Package for Seamless Multicollinearity Management


<!-- Development badges 

[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)
[![Devel-version](https://img.shields.io/badge/devel%20version-1.0.1-blue.svg)](https://github.com/blasbenito/collinear)

<!-- badges: start -->

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10039489.svg)](https://doi.org/10.5281/zenodo.10039489)
[![CRAN status](https://www.r-pkg.org/badges/version/collinear)](https://cran.r-project.org/package=collinear)
[![CRAN\_Download\_Badge](http://cranlogs.r-pkg.org/badges/grand-total/collinear)](https://CRAN.R-project.org/package=collinear)
[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

## Warning

Version 2.0.0 of `collinear` includes changes that may disrupt existing workflows, and results from previous versions may not be reproducible due to enhancements in the automated selection algorithms. Please refer to the Changelog for details.

## Summary

The `collinear` package combines four methods for easy management of multicollinearity in modelling data frames with numeric and categorical variables:

- **Target Encoding**: Transforms categorical predictors to numeric using a numeric response as reference.
- **Preference Order**: Ranks predictors by their association with a response variable to preserve important ones in multicollinearity filtering.
- **Pairwise Correlation Filtering**: Automated multicollinearity filtering of numeric and categorical predictors based on pairwise correlations.
- **Variance Inflation Factor Filtering**: Automated multicollinearity filtering of numeric predictors based on Variance Inflation Factors.

These methods are combined in the function `collinear()`, and as individual functions in `target_encoding_lab()`, `preference_order()`, `cor_select()`, and `vif_select()`.

## Citation

If you find this package useful, please cite:

*Blas M. Benito (2024). collinear: R Package for Seamless Multicollinearity Management. Version 1.2.0. doi: 10.5281/zenodo.10039489*

## Main Improvements in Version 2.0.0

1. **Expanded Functionality**: Functions `collinear()` and `preference_order()` support both categorical and numeric responses and predictors, and can handle several responses at once.
2. **Robust Selection Algorithms**: Enhanced selection in `vif_select()` and `cor_select()`.
3. **Refined `preference_order()`**: New functions to compute association between response and predictors covering most use-cases, and automated function selection depending on data features.
4. **Simplified Target Encoding**: Streamlined and parallelized for better efficiency, and new default is "loo" (leave-one-out).
5. **Parallelization and Progress Bars**: Utilizes `future` and `progressr` for enhanced performance and user experience.


## Install

The package `collinear` can be installed from CRAN.

```{r, eval = FALSE}
install.packages("collinear")
```


The development version can be installed from GitHub.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "development"
  )
```


Previous versions are in the “archive_xxx” branches of the GitHub
repository.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "archive_v1.1.1"
  )
```

## Required Packages

The packages required for this tutorial are:

```{r packages, message = FALSE, warning = FALSE, include = FALSE}
library(collinear)
library(dplyr)
library(future)
library(parallelly)
library(kableExtra)
```

## Parallelization Setup and Progress Bars

Most functions in the package now accept a parallelization setup via `future::plan()` and progress bars via `progressr::handlers()`. However, progress bars are ignored in this tutorial because they don't work in Rmarkdown.

```{r parallelization}
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#progressr::handlers(global = TRUE)
```


## Example Data

The data frame `vi` consists of `r nrow(vi)` rows, `r ncol(vi)` columns, and `r sum(is.na(vi))` NA values. It contains several numeric and categorical responses and predictors.

The response columns, all derived from the same data, have descriptive names: `vi_numeric`, `vi_counts` (integers), `vi_binomial` (1s and 0s), `vi_categorical` (five categories), and `vi_factor` (factor version of the previous one).

Predictor names are grouped in character vectors: `vi_predictors_numeric` (49 numeric and integer predictors), `vi_predictors_categorical` (12 character and factor predictors), and `vi_predictors` containing them all.

The code below makes it smaller to accelerate this tutorial.

```{r make_vi_smaller}
set.seed(1)
df <- dplyr::slice_sample(
  vi, 
  n = 5000
  )
```

## Using `collinear()`

The function `collinear()` provides access to the key functionalities of this package. The code below shows a call to `collinear()` with two responses and mixed predictor types.

```{r collinear_demo}
selection <- collinear::collinear(
  df = df,
  response = c(
    "vi_numeric",
    "vi_categorical"
    ),
  predictors = vi_predictors,
  quiet = FALSE
)
```

The function returns a named list of vectors with predictor names when more than one response is provided, and a character vector otherwise.  

```{r}
selection
```

If `response` is provided, and `preference_order` is provided or has the value "auto", then `collinear()` returns the variable selection in order of preference. Otherwise, the predictors are ordered from lower to higher multicollinearity.

### How It Works

The table below shows the functions called directly by `collinear()`, their functionality, data requirements, and how to disable them.

| **Function**           | **Functionality**                            | **Requirements**                  | **Disabled**                              |
|------------------------|----------------------------------------------|-----------------------------------|------------------------------------------|
| `target_encoding_lab()`| categorical predictors <br> to numeric       | - numeric `response` <br> - categorical `predictors` | - `response = NULL` <br> - `encoding_method = NULL` |
| `preference_order()`   | rank and preserve <br> important predictors  | any `response`                    | - `response = NULL` <br> - `preference_order = NULL` |
| `cor_select()`         | reduce <br> pairwise correlation             | any `predictors`                  | `cor_max = NULL`                         |
| `vif_select()`         | reduce <br> variance inflation               | numeric `predictors`              | `vif_max = NULL`                         |

The next sections follow the processing of the responses "vi_numeric" and "vi_character" through these functions.

#### `target_encoding_lab()`

Target-encoding requires a numeric `response` to transform categorical `predictors` to numeric. This transformation enables the application of the same multicollinearity filtering (and modelling) methods to categorical and numeric predictors.

In `collinear()`, this functionality is controlled by the argument `encoding_method`, which disables the functionality entirely when `NULL`, or defines the encoding method as "loo" (leave-one-out, default method), "mean", or "rank".

There is a lengthier article about target-encoding [here](https://www.blasbenito.com/post/target-encoding/), let me show you a brief example of how it works. The example data frame below has two levels of the categorical predictor "koppen_zone" and the response "vi_numeric".

```{r, echo = FALSE}
df_toy <- df |> 
  dplyr::select(vi_numeric, koppen_zone) |> 
  dplyr::filter(koppen_zone %in% c("Af", "BSh")) |> 
  dplyr::group_by(koppen_zone) |> 
  dplyr::slice_head(n = 5) |> 
  dplyr::ungroup()

kableExtra::kbl(df_toy) |> 
  kableExtra::kable_styling()
```

When introducing this data frame into `target_encoding_lab()` with the method "loo", it is first grouped by "koppen_zone" levels, and then each case  is encoded as the average of `response` across all other cases within the same level. 

```{r, echo = FALSE}
df_toy <- collinear::target_encoding_lab(
  df = df_toy,
  response = "vi_numeric",
  predictors = "koppen_zone",
  method = "loo",
  overwrite = TRUE,
  quiet = TRUE
)
```

The result shows "koppen_zone" encoded as numeric and ready for a multicollinearity analysis.

```{r, echo = FALSE}
kableExtra::kbl(df_toy) |> 
  kableExtra::kable_styling()
```

Due to the requirement for a numeric `response`, target encoding is only applied for the `response` "vi_numeric" as follows:

```{r}
df_vi_numeric <- collinear::target_encoding_lab(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors,
  method = "loo",
  overwrite = TRUE,
  quiet = TRUE
)
```

This operation results in zero categorical predictors in the vector `vi_predictors`:

```{r}
collinear::identify_predictors_categorical(
  df = df_vi_numeric,
  predictors = vi_predictors
)
```

On the other hand, target encoding is skipped for the categorical `response` "vi_categorical", resulting in 12 categorical predictors.

```{r}
df_vi_categorical <- df

collinear::identify_predictors_categorical(
  df = df_vi_categorical,
  predictors = vi_predictors
)
```

#### `preference_order()`

The functions `collinear()`, `cor_select()`, and `vif_select()` are designed to preserve as many "relevant" predictors as possible during the multicollinearity filtering. This functionality is controlled by the argument `preference_order`. This argument accepts two different inputs, a custom character vector or a data frame.

**Custom character vector**

Vector with `predictors` names in order of interest for the analyst. The example below shows a hypothetical case focused on preserving soil temperature variables over all others.
  
```{r}
selection_from_vector <- collinear::collinear(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors_numeric,
  preference_order = c(
    "soil_temperature_mean",
    "soil_temperature_range",
    "soil_temperature_min",
    "soil_temperature_max"
  ),
  quiet = TRUE
)

selection_from_vector
```
**Data frame**

Must have the column "predictor", and should be ordered from higher to lower quantitative preference. 

The function `preference_order()` generates this data frame by computing the univariate association between each predictor and the `response` using a given `f` function:
  
```{r}
preference_df <- collinear::preference_order(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors_numeric,
  f = f_r2_pearson,
  quiet = TRUE
)
```


```{r, echo = FALSE}
kableExtra::kbl(preference_df) |> 
  kableExtra::kable_styling()
```
The resulting data frame can be plugged into the `preference_order` argument of `collinear()`:

```{r}
selection_from_df <- collinear::collinear(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors_numeric,
  preference_order = preference_df,
  quiet = TRUE
)

selection_from_df
```
But `collinear()` can also compute preference order on its own to obtain the same result:

```{r}
selection_auto <- collinear::collinear(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors_numeric,
  preference_order = "auto",
  preference_f = "auto",
  quiet = TRUE
)

selection_auto
```
The argument `f` of `preference_order()` (named `preference_f` in `collinear()`) either receives a function name, or the string "auto".

**Function name**

The function `f_functions()` returns a data frame with the details of all `f` functions.

```{r, eval = FALSE}
collinear::f_functions()
```

```{r, echo = FALSE}
collinear::f_functions() |> 
  kableExtra::kbl() |> 
  kableExtra::kable_styling()
```
All these functions take a data frame named `df` with the columns "x" and "y" as input, so preparing a custom one is a simple task. For example, the one below returns the R-squared of a linear model between the predictor and the response. One just have to take in mind that `preference_order()` will rank the results of such function from higher to lower values.

```{r}
f_lm <- function(df){
  summary(lm(y ~ x, data = df))$r.squared
}

preference_lm <- collinear::preference_order(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors_numeric,
  f = f_lm,
  quiet = TRUE
)
```

```{r, echo = FALSE}
kableExtra::kbl(preference_lm) |> 
  kableExtra::kable_styling()
```





, and it does not require a numeric response or predictors to do so:

```{r}
selection_categorical <- collinear::collinear(
  df = df,
  response = "vi_categorical",
  predictors = vi_predictors_categorical,
  preference_order = "auto",
  preference_f = "auto",
  quiet = TRUE
)

selection_categorical
```




#### `cor_select()` 

#### `vif_select()` 


If you got here, thank you for your interest in `collinear`. I hope you can find it useful!

Blas M. Benito, PhD
