---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
# options(tibble.print_min = 5, tibble.print_max = 5)
```


# collinear: R Package for Seamless Multicollinearity Management


<!-- Development badges 

[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)
[![Devel-version](https://img.shields.io/badge/devel%20version-1.0.1-blue.svg)](https://github.com/blasbenito/collinear)

<!-- badges: start -->

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10039489.svg)](https://doi.org/10.5281/zenodo.10039489)
[![CRAN status](https://www.r-pkg.org/badges/version/collinear)](https://cran.r-project.org/package=collinear)
[![CRAN\_Download\_Badge](http://cranlogs.r-pkg.org/badges/grand-total/collinear)](https://CRAN.R-project.org/package=collinear)
[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

## Warning

Version 2.0.0 of `collinear` includes changes that may disrupt existing workflows, and results from previous versions may not be reproducible due to enhancements in the automated selection algorithms. Please refer to the Changelog for details.

## Summary

The `collinear` package combines four methods for easy management of multicollinearity in modelling data frames with numeric and categorical variables:

- **Target Encoding**: Transforms categorical predictors to numeric using a numeric response as reference.
- **Preference Order**: Ranks predictors by their association with a response variable to preserve important ones in multicollinearity filtering.
- **Pairwise Correlation Filtering**: Automated multicollinearity filtering of numeric and categorical predictors based on pairwise correlations.
- **Variance Inflation Factor Filtering**: Automated multicollinearity filtering of numeric predictors based on Variance Inflation Factors.

These methods are combined in the function `collinear()`, and as individual functions in `target_encoding_lab()`, `preference_order()`, `cor_select()`, and `vif_select()`.

## Citation

If you find this package useful, please cite:

*Blas M. Benito (2024). collinear: R Package for Seamless Multicollinearity Management. Version 1.2.0. doi: 10.5281/zenodo.10039489*

## Main Improvements in Version 2.0.0

1. **Expanded Functionality**: Functions `collinear()` and `preference_order()` support both categorical and numeric responses and predictors, and can handle several responses at once.
2. **Robust Selection Algorithms**: Enhanced selection in `vif_select()` and `cor_select()`.
3. **Refined `preference_order()`**: New functions to compute association between response and predictors covering most use-cases, and automated function selection depending on data features.
4. **Simplified Target Encoding**: Streamlined and parallelized for better efficiency, and new default is "loo" (leave-one-out).
5. **Parallelization and Progress Bars**: Utilizes `future` and `progressr` for enhanced performance and user experience.


## Install

The package `collinear` can be installed from CRAN.

```{r, eval = FALSE}
install.packages("collinear")
```


The development version can be installed from GitHub.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "development"
  )
```


Previous versions are in the “archive_xxx” branches of the GitHub
repository.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "archive_v1.1.1"
  )
```

## Required Packages

The packages required for this tutorial are:

```{r packages, message = FALSE, warning = FALSE, include = FALSE}
library(collinear)
library(dplyr)
library(future)
library(parallelly)
library(kableExtra)
```

## Parallelization Setup and Progress Bars

Most functions in the package now accept a parallelization setup via `future::plan()` and progress bars via `progressr::handlers()`. However, progress bars are ignored in this tutorial because they don't work in Rmarkdown.

```{r parallelization}
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#progressr::handlers(global = TRUE)
```


## Example Data

The data frame `vi` consists of `r nrow(vi)` rows, `r ncol(vi)` columns, and `r sum(is.na(vi))` NA values. It contains several numeric and categorical responses and predictors.

The response columns, all derived from the same data, have descriptive names: `vi_numeric`, `vi_counts` (integers), `vi_binomial` (1s and 0s), `vi_categorical` (five categories), and `vi_factor` (factor version of the previous one).

Predictor names are grouped in character vectors: `vi_predictors_numeric` (49 numeric and integer predictors), `vi_predictors_categorical` (12 character and factor predictors), and `vi_predictors` containing them all.

The code below makes it smaller to accelerate this tutorial.

```{r make_vi_smaller}
set.seed(1)
df <- dplyr::slice_sample(
  vi, 
  n = 5000
  )
```

## Using `collinear()`

The function `collinear()` provides access to the key functionalities of this package. The code below shows a call to `collinear()` with two responses and mixed predictor types.

```{r collinear_demo}
selection <- collinear::collinear(
  df = df,
  response = c(
    "vi_numeric",
    "vi_categorical"
    ),
  predictors = vi_predictors,
  quiet = FALSE
)
```

The function returns a named list of vectors with predictor names when more than one response is provided, and a character vector otherwise.  

```{r}
selection
```

If `response` is provided, and `preference_order` is provided or has the value "auto", then `collinear()` returns the variable selection in order of preference. Otherwise, the predictors are ordered from lower to higher multicollinearity.

### How It Works

The table below shows the functions called directly by `collinear()`, their functionality, data requirements, and how to disable them.

| **Function**           | **Functionality**                            | **Requirements**                  | **Disabled**                              |
|------------------------|----------------------------------------------|-----------------------------------|------------------------------------------|
| `target_encoding_lab()`| categorical predictors <br> to numeric       | - numeric `response` <br> - categorical `predictors` | - `response = NULL` <br> - `encoding_method = NULL` |
| `preference_order()`   | rank and preserve <br> important predictors  | any `response`                    | - `response = NULL` <br> - `preference_order = NULL` |
| `cor_select()`         | reduce <br> pairwise correlation             | any `predictors`                  | `cor_max = NULL`                         |
| `vif_select()`         | reduce <br> variance inflation               | numeric `predictors`              | `vif_max = NULL`                         |

The next sections follow the processing of the responses "vi_numeric" and "vi_character" through these functions.

#### `target_encoding_lab()`

This function requires a numeric `response` to transform categorical `predictors` to numeric. This transformation allows applying the same multicollinearity filtering methods to categorical and numeric predictors. 

In `collinear()`, this functionality is controlled by the argument `encoding_method`, which defines what method will be used for the target encoding, or disables the functionality entirely when `NULL`.

The code chunk below creates an example data frame with two levels of the categorical predictor "koppen_zone" and the response "vi_numeric".

```{r}
df_toy <- df |> 
  dplyr::select(vi_numeric, koppen_zone) |> 
  dplyr::filter(koppen_zone %in% c("Af", "BSh")) |> 
  dplyr::group_by(koppen_zone) |> 
  dplyr::slice_head(n = 5) |> 
  dplyr::ungroup()
```

```{r, echo = FALSE}
kableExtra::kbl(df_toy) |> 
  kableExtra::kable_styling()
```
When introducing this data frame into `target_encoding_lab()` with the method "loo" (leave-one-out), the data frame is grouped by the levels of "koppen_zone", and each value in a group is encoded as the average of `response` across all other group cases. 

The methods "mean" and "rank" are also available through the argument `encoding_method` of `collinear()`. Please check the documentation of `target_encoding_lab()` for further details.

```{r}
df_toy <- collinear::target_encoding_lab(
  df = df_toy,
  response = "vi_numeric",
  predictors = "koppen_zone",
  method = "loo",
  overwrite = TRUE,
  quiet = TRUE
)
```

The function returns a data frame with the predictor "koppen_zone" encoded as numeric, and ready for a multicollinearity analysis along with other numeric predictors.

```{r, echo = FALSE}
kableExtra::kbl(df_toy) |> 
  kableExtra::kable_styling()
```
In the example call to `collinear()` shown at the beginning of this section, target encoding is executed for "vi_numeric" only, as follows:

```{r}
df_vi_numeric <- collinear::target_encoding_lab(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors,
  method = "loo",
  overwrite = TRUE,
  quiet = TRUE
)
```

This operation results in zero categorical predictors in the vector `vi_predictors`:

```{r}
collinear::identify_predictors_categorical(
  df = df_vi_numeric,
  predictors = vi_predictors
)
```

On the other hand, target encoding is skipped for the categorical `response` "vi_categorical", resulting in 12 categorical predictors that will require a different handling than the numeric ones in the next steps.

```{r}
df_vi_categorical <- df

collinear::identify_predictors_categorical(
  df = df_vi_categorical,
  predictors = vi_predictors
)
```

#### `preference_order()`

This function ranks predictors by the strength of their association to the response to help preserve a relevant subset during multicollinearity filtering.

Three arguments control this feature in `collinear()`:

  + `preference_order`: Either a pre-computed or custom ranking of predictors, or "auto" to let `collinear()` handle it internally.
  + `preference_f`: Function to compute the association between `response` and `predictors` (check `help(preference_order)`, which lists the available options). If "auto" (default), then `f_auto()` selects one based on the data features.
  + `preference_warn_limit`: If not `NULL`, this feature helps identifying predictors that are suspiciously similar to the response.
  
Preference order is computed as follows for the `response` "vi_numeric":

```{r}
preference_vi_numeric <- collinear::preference_order(
  df = df_vi_numeric,
  response = "vi_numeric",
  predictors = vi_predictors,
  f = "auto",
  warn_limit = 0.95,
  quiet = FALSE
)
```
When all variables are numeric, `f_auto()` selects `f_r2_pearson()` to compute the Pearson correlation between `response` and `predictors`. 

The function returns the data frame shown below, ordered by the value of the column "preference".

```{r, echo = FALSE}
kableExtra::kbl(preference_vi_numeric) |> 
  kableExtra::kable_styling()
```

The order of the column "predictor" matches the output of `collinear()` for the response "vi_numeric".

```{r}
selection$vi_numeric
```
This data frame resulting from `preference_order()` can be used as input for the `preference_order` argument of `collinear()`.

If the choice made by `f_auto()` is deemed inappropriate, there are other options.

```{r}
preference_gam <- collinear::preference_order(
  df = df_vi_numeric,
  response = "vi_numeric",
  predictors = vi_predictors,
  f = f_r2_gam_gaussian,
  quiet = TRUE
)
```

```{r, echo = FALSE}
kableExtra::kbl(preference_gam) |> 
  kableExtra::kable_styling()
```
The argument `preference_order` of `collinear()` also accepts a user-defined vector of preference. The example below shifts the focus of the multicollinearity analysis towards several soil-related predictors.

```{r}
collinear::collinear(
  df = df,
  response = "vi_numeric",
  predictors = vi_predictors,
  preference_order = c(
    "soil_ph",
    "soil_clay",
    "soil_silt",
    "soil_sand",
    "soil_nitrogen",
    "soil_soc",
    "soil_temperature_mean"
  ),
  quiet = TRUE,
  cor_max = 0.5,
  vif_max = 2.5
)
```
In this case, `collinear()` passes all predictors not in the custom vector to `preference_order()`, and appends the resulting ranking to the custom vector.

#### `cor_select()` 

#### `vif_select()` 


If you got here, thank you for your interest in `collinear`. I hope you can find it useful!

Blas M. Benito, PhD
