---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
# options(tibble.print_min = 5, tibble.print_max = 5)
```


# collinear: R Package for Seamless Multicollinearity Management


<!-- Development badges 

[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)
[![Devel-version](https://img.shields.io/badge/devel%20version-1.0.1-blue.svg)](https://github.com/blasbenito/collinear)

<!-- badges: start -->

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10039489.svg)](https://doi.org/10.5281/zenodo.10039489)
[![CRAN status](https://www.r-pkg.org/badges/version/collinear)](https://cran.r-project.org/package=collinear)
[![CRAN\_Download\_Badge](http://cranlogs.r-pkg.org/badges/grand-total/collinear)](https://CRAN.R-project.org/package=collinear)
[![R-CMD-check](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/collinear/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

## Warning

Version 2.0.0 of `collinear` includes changes that may disrupt existing workflows, and results from previous versions may not be reproducible due to enhancements in the automated selection algorithms. Please refer to the Changelog for details.

## Summary

The `collinear` package combines four methods for easy management of multicollinearity in modelling data frames with numeric and categorical variables:

- **Target Encoding**: Transforms categorical predictors to numeric using a numeric response as reference.
- **Preference Order**: Ranks predictors by their association with a response variable to preserve important ones in multicollinearity filtering.
- **Pairwise Correlation Filtering**: Automated multicollinearity filtering of numeric and categorical predictors based on pairwise correlations.
- **Variance Inflation Factor Filtering**: Automated multicollinearity filtering of numeric predictors based on Variance Inflation Factors.

These methods are combined in the function `collinear()`, and as individual functions in `target_encoding_lab()`, `preference_order()`, `cor_select()`, and `vif_select()`.

## Citation

If you find this package useful, please cite:

*Blas M. Benito (2024). collinear: R Package for Seamless Multicollinearity Management. Version 1.2.0. doi: 10.5281/zenodo.10039489*

## Main Improvements in Version 2.0.0

1. **Expanded Functionality**: Functions `collinear()` and `preference_order()` support both categorical and numeric responses and predictors, and can handle several responses at once.
2. **Robust Selection Algorithms**: Enhanced selection in `vif_select()` and `cor_select()`.
3. **Refined `preference_order()`**: New functions to compute association between response and predictors covering most use-cases, and automated function selection depending on data features.
4. **Simplified Target Encoding**: Streamlined and parallelized for better efficiency, and new default is "loo" (leave-one-out).
5. **Parallelization and Progress Bars**: Utilizes `future` and `progressr` for enhanced performance and user experience.


## Install

The package `collinear` can be installed from CRAN.

```{r}
install.packages("collinear")
```


The development version can be installed from GitHub.

```{r}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "development"
  )
```


Previous versions are in the “archive_xxx” branches of the GitHub
repository.

```{r}
remotes::install_github(
  repo = "blasbenito/collinear", 
  ref = "archive_v1.1.1"
  )
```

## Required Packages

The packages required for this tutorial are:

```{r packages, message = FALSE, warning = FALSE, include = FALSE}
library(collinear)
library(dplyr)
library(future)
library(parallelly)
```

## Parallelization Setup

Most functions in the package now accept a parallelization setup and progress bars:

```{r parallelization}
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#NOTE: progress bars do not work in .Rmd files
#progressr::handlers(global = TRUE)
```


## Example Data

The data frame `vi` consists of 30,000 rows and 67 columns, featuring various response types and a mix of numeric and categorical predictors. The code below makes it smaller to accelerate this tutorial.

```{r make_vi_smaller}
set.seed(1)
df <- dplyr::slice_sample(
  vi, 
  n = 5000
  )
```


The response columns, all derived from the same data, have descriptive names: `vi_numeric`, `vi_counts` (integers), `vi_binomial` (1s and 0s), `vi_categorical` (five categories), and `vi_factor` (factor version of the previous one).

Predictor names are grouped in character vectors: `vi_predictors_numeric` (49 numeric and integer predictors), `vi_predictors_categorical` (12 character and factor predictors), and `vi_predictors` containing them all.

The plot below shows the hierarchical clustering of the correlation matrix between predictors in `vi`.

```{r figure_1, fig.width=12, fig.height=5, warning = FALSE, fig.cap="Figure 1: groups of predictors in the data frame vi with a correlation higher than 0.5."}
df_clusters <- collinear::cor_clusters(
  df = df, 
  predictors = vi_predictors, 
  cor_max = 0.5,
  plot = TRUE
  )
```

## Using `collinear()`

The function `collinear()` provides access to the key functionalities of this package. The table below shows these functionalities, the functions implementing them, their requirements, and how to disable them within `collinear()`.

| **Functionality** |**Function**|**Requirements** | **Disable** |
|-------------------|------------|-----------------|--------------|
| categorical predictors <br> to numeric|`target_encoding_lab()` | - numeric `response` <br> - categorical `predictors` | - `response = NULL` <br> - `encoding_method = NULL`|
| rank and preserve <br> important predictors | `preference_order()` | any `response` | - `response = NULL` <br> - `preference_order = NULL`|
| reduce <br> pairwise correlation | `cor_select()` | any `predictors` | `cor_max = NULL` |
| reduce <br> variance inflation | `vif_select()`| numeric `predictors` | `vif_max = NULL` |

The code below shows a full call to `collinear()` with two responses and mixed predictor types.

```{r}
selection <- collinear::collinear(
  df = df,
  response = c(
    "vi_numeric",
    "vi_categorical"
    ),
  predictors = vi_predictors,
  encoding_method = "loo",
  preference_order = "auto",
  preference_f = "auto",
  preference_warn_limit = 0.8,
  cor_method = "pearson",
  cor_max = 0.75,
  vif_max = 5,
  quiet = FALSE
)
```

The function returns a vector of predictor names.

```{r}
selection
```





The function returns a vector of predictor ordered by preference with a pairwise correlation lower than 0.75 and a VIF lower than 5. The code below shows the pairwise correlations and the VIF scores of the selected predictors.

```{r}
#categorical predictors require target-encoding to mimic behavior of collinear()
df <- target_encoding_lab(
  df = vi,
  response = "vi_numeric",
  predictors = selected_predictors,
  encoding_method = "mean",
  replace = TRUE
)

#pairwise correlation data frame
cor_df(
  df = df,
  predictors = selected_predictors
)

#vif data frame
vif_df(
  df = df, #data frame with encoded categoricals
  predictors = selected_predictors
) |> 
  dplyr::arrange(
    dplyr::desc(vif)
  )
```
## `collinear()` Step by Step

This section describes the steps `collinear()` follows to deliver its results.

### Data Validation

There are several functions called within `collinear()` to validate the input data:

  + `validate_df()`: checks basic details about the input data frame

#### Preference order

The argument `preference_order` gives the user some control on what predictors should be removed first and what predictors should be kept during the multicollinearity filtering. This argument accepts a vector of predictor names in the order of interest, or the result of the function `preference_order()`, which allows to define preference order following a quantitative criteria. 

##### Manual preference order

Let's start with the former option. Below, the argument `preference_order` names several predictors that are of importance for a hypothetical analysis. The `predictors` not in `preference_order` are ranked by the absolute sum of their correlations with other predictors during the pairwise correlation filtering, and by their VIF during the VIF-based filtering.

```{r}
selected_predictors <- cor_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = c(
    "soil_temperature_mean",
    "soil_temperature_max",
    "soil_type"
  )
)

selected_predictors
```

Notice that in the output, two of the variables in `preference_order` are selected ("soil_temperature_mean" and "soil_type"), but one was removed ("soil_temperature_max"). This happens because at some point in the selection, the VIF of "soil_temperature_mean" and "soil_temperature_max" was higher than `vif_max`, and the one with lower preference was removed.

##### Quantitative preference order

The function `preference_order()` requires the `response` argument, and takes a function `f` that returns a value of association between the response and any predictor. This value is then located in the "preference" column of the function's output.

```{r}
#parallelization setup
# future::plan(
#   future::multisession,
#   workers = 2 #set to parallelly::availableWorkers() - 1
# )

#progress bar
#progressr::handlers(global = TRUE)

#compute preference order
preference_rsquared <- preference_order(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  f = f_rsquared,
)

preference_rsquared
```

The result of `preference_order()` can be plugged right away into the `preference_order` argument of collinear.

```{r}
selected_predictors <- collinear(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = preference_rsquared
)
selected_predictors
```

This variable selection satisfies three conditions at once: maximum correlation between each predictor and the response, maximum pairwise correlation, and maximum VIF.

The `f` argument used by default is the function `f_rsquared()`, that returns the R-squared between the response and any predictor.

```{r}
f_rsquared(
  x = "growing_season_length",
  y = "vi_mean",
  df = vi
)
```

There are several other `f` functions implemented:

  + `f_gam_deviance()`: returns the explained deviance of a univariate GAM model between the response and each predictor, fitted with the function `mgcv::gam()`. Only if the R package `mgcv` is installed in the system.
  + `f_rf_rsquared()` (also named `f_rf_deviance()`): returns the explained deviance of a univariate Random Forest model between the response and each predictor, fitted with the function `ranger::ranger()`. Only if the R package `ranger` is installed in the system.
  + `f_logistic_auc_balanced()` and `f_logistic_auc_unbalanced()`: return the area under the ROC curve of univariate binomial GLM between a binary response of 1s and 0s and a numeric predictor. The former assumes the response is balanced, while the latter applies case weights to mitigate unbalances.
  + `f_gam_auc_balanced()` and `f_gam_auc_unbalanced()`: return the area under the ROC curve of univariate binomial GAM between a binary response of 1s and 0s and a numeric predictor. The former assumes the response is balanced, while the latter applies case weights to mitigate unbalances.
  + `f_rf_auc_balanced()` and `f_rf_auc_unbalanced()`: return the area under the ROC curve of univariate random forest models between a binary response of 1s and 0s and a numeric predictor. The former assumes the response is balanced, while the latter applies case weights to mitigate unbalances.
  
```{r, eval = FALSE}
#example of preference order for a binary variable

#the binary variable
table(vi$vi_binary)

#computation of preference order with 
preference_auc <- preference_order(
  df = vi,
  response = "vi_binary",
  predictors = vi_predictors,
  f = f_gam_auc_unbalanced
)

preference_auc
```
  
Custom functions created by the user are also accepted as input, as long as they have the `x`, `y`, and `df` arguments, and they return a single numeric value.

### `cor_select()` and `vif_select()`

The functions `cor_select()` and `vif_select()`, called within `collinear()`, perform the pairwise correlation filtering, and the VIF-based filtering. The main difference between them is that `cor_select()` can handle categorical predictors even when the `response` is omitted, while `vif_select()` ignores them entirely in such case.

```{r}
selected_predictors_cor <- cor_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = preference_rsquared
)

selected_predictors_vif <- vif_select(
  df = vi,
  response = "vi_mean",
  predictors = vi_predictors,
  preference_order = preference_rsquared
)
```

```{r}
selected_predictors_cor
```

```{r}
selected_predictors_vif
```

### `target_encoding_lab()`

The function `target_encoding_lab()` is used within all other functions in the package to encode categorical variables as numeric. It implements four target encoding methods:

 + "mean" (in `target_encoding_mean()`): replaces each category with the mean of the response across the category. White noise can be added to this option to increase data variability.
 + "loo" (in `target_encoding_loo()`): replaces each value in a category with the mean of the response across all the other cases within the category. White noise can be added to this option to increase data variability.
 + "rank" (in `target_encoding_rank()`): replaces each category with the rank of the mean of the response across the category.
 
The method "mean" is used as default throughout all functions in the package, but can be changed via the argument `encoding_method`.
 
Below we use all methods to generate different numeric encodings for the categorical variable "koppen_zone".

```{r}
df <- target_encoding_lab(
  df = vi,
  response = "vi_mean",
  predictors = "koppen_zone",
  encoding_methods = c(
    "mean",
    "loo",
    "rank"
  ),
  seed = 1,
  white_noise = c(0, 0.01, 0.1),
  quiet = FALSE
)
```

The relationship between these encoded versions of "koppen_zone" and the response are shown below.

```{r, fig.width=9, fig.height=9, eval = FALSE}
#get names of encoded variables
koppen_zone_encoded <- grep(
  pattern = "*__encoded*",
  x = colnames(df),
  value = TRUE
)

#record the user's graphical parameters
user.par <- par(no.readonly = TRUE)

#modify graphical parameters for the plot
par(mfrow = c(4, 3))

#plot target encoding
x <- lapply(
  X = koppen_zone_encoded,
  FUN = function(x) plot(
    x = df[[x]],
    y = df$vi_mean,
    xlab = x,
    ylab = "vi_mean",
    cex = 0.5
    )
)

#reset the user's graphical parameters
par(user.par)
```

The function implementing each method can be used directly as well. The example below shows the "mean" method with the option `replace = FALSE`, which replaces the categorical values with the numeric ones in the output data frame.

```{r}
head(vi[, c("vi_mean", "koppen_zone")], n = 10)
```


```{r}
df <- target_encoding_mean(
  df = vi,
  response = "vi_mean",
  predictor = "koppen_zone",
  replace = TRUE
)

head(df[, c("vi_mean", "koppen_zone")], n = 10)
```

If you got here, thank you for your interest in `collinear`. I hope you can find it useful!

Blas M. Benito, PhD
