---
title: "Development Tracking"
output: html_document
---

# Development Tools

```{r, message = FALSE, warning = FALSE}
library(testthat)
library(devtools)
library(covr)
library(codetools)
library(pkgnet)
```

Load package functions

```{r}
devtools::load_all()
```

# Internal Dependencies

```{r, eval = FALSE}
out <- pkgnet::CreatePackageReport(
  "collinear", 
  report_path = "report.html"
  )
```

# List Functions

```{r, eval = FALSE}
#copy-paste output below
message(
  paste(
    ls(getNamespace("collinear"), all.names = TRUE), 
    collapse = "\n")
  )
```

```         
DONE case_weights
DONE class.collinear_selection
DONE collinear_stats
DONE collinear
DONE collinear_select
DONE collinear_auto
DONE cor_clusters
DONE cor_cramer
DONE cor_df
DONE cor_matrix
DONE cor_select
DONE cor_stats
DONE drop_geometry_column
DONE f_binomial_gam
DONE f_binomial_glm
DONE f_binomial_rf
DONE f_auto
DONE f_auto_rules
DONE f_functions
DONE f_numeric_gam
DONE f_count_gam
DONE f_count_rf
DONE f_count_glm
DONE f_numeric_rf
DONE f_numeric_glm
DONE f_numeric_gam
DONE f_categorical_rf
DONE identify_valid_variables
DONE identify_categorical_variables
DONE identify_logical_variables
DONE identify_numeric_variables
DONE identify_zero_variance_variables
DONE identify_response_type
DONE model_formula
DONE score_auc
DONE score_r2
DONE score_cramer
DONE preference_order
DONE print.collinear_output
DONE print.collinear_selection
DONE summary.collinear_output
DONE summary.collinear_selection
DONE target_encoding_lab
DONE target_encoding_loo
DONE target_encoding_mean
DONE target_encoding_rank
DONE validate_arg_df
DONE validate_arg_df_not_null
DONE validate_arg_encoding_method
DONE validate_arg_f
DONE validate_arg_function_name
DONE validate_arg_predictors
DONE validate_arg_preference_order
DONE validate_arg_quiet
DONE validate_arg_responses
DONE validate_arg_max_cor
DONE validate_arg_max_vif
DONE vif
DONE vif_df
DONE vif_stats
DONE vif_select
```

# TODO

- Modify `collinear()` because `vif_select()` accepts non-numeric predictors, as it uses `cor_matrix()` behind the scenes.

# Function by Function

## DONE validate_arg_function_name

Depends: none

```{r}
f_name <- "validate_arg_function_name"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_quiet

Depends:

-   validate_arg_function_name

```{r}
f_name <- "validate_arg_quiet"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_df_not_null

Depends:

-   validate_arg_function_name

```{r}
f_name <- "validate_arg_df_not_null"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_predictors

Depends:

- validate_arg_df_not_null  
- validate_arg_function_name

```{r}
f_name <- "validate_arg_predictors"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_responses

Depends:

-   validate_arg_df_not_null
-   validate_arg_function_name

```{r}
f_name <- "validate_arg_responses"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE drop_geometry_column

Depends:

-   validate_arg_function_name

```{r}
f_name <- "drop_geometry_column"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE identify_zero_variance_variables

Depends:

 - validate_arg_df_not_null  
 - validate_arg_function_name
 - validate_arg_predictors  
 - validate_arg_responses 
 - validate_arg_quiet

```{r}
f_name <- "identify_zero_variance_variables"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```



## DONE identify_numeric_variables

Depends:

-   identify_zero_variance_variables
-   validate_arg_df_not_null        
-   validate_arg_function_name      
-   validate_arg_predictors
-   validate_arg_responses
-   validate_arg_quiet

```{r}
f_name <- "identify_numeric_variables"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE identify_logical_variables

Depends:

-   validate_arg_df_not_null 
-   validate_arg_function_name
-   validate_arg_predictors 
-   validate_arg_responses
-   validate_arg_quiet

```{r}
f_name <- "identify_logical_variables"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE identify_categorical_variables

Depends:

-   validate_arg_df_not_null  
-   validate_arg_function_name
-   validate_arg_predictors
-   validate_arg_responses
-   validate_arg_quiet

```{r}
f_name <- "identify_categorical_variables"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE identify_valid_variables

Depends:

-  identify_categorical_variables
-  identify_logical_variables    
-  identify_numeric_variables
-  validate_arg_df_not_null       
-  validate_arg_function_name     
-  validate_arg_predictors

```{r}
f_name <- "identify_valid_variables"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```


## DONE identify_response_type

Depends:

- validate_arg_df           
- validate_arg_df_not_null 
- validate_arg_function_name
- validate_arg_quiet        
- validate_arg_responses 

```{r}
f_name <- "identify_response_type"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```



## DONE validate_arg_df

Depends:

-   drop_geometry_column
-   identify_valid_variables
-   validate_arg_df_not_null 
-   validate_arg_function_name
-   validate_arg_quiet 

```{r}
f_name <- "validate_arg_df"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```






# TARGET ENCODING

## DONE validate_arg_encoding_method

Depends:

-   validate_arg_function_name

```{r}
f_name <- "validate_arg_encoding_method"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE target_encoding_mean

Depends:

-   none

```{r}
f_name <- "target_encoding_mean"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE target_encoding_rank

Depends:

-   none

```{r}
f_name <- "target_encoding_rank"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE target_encoding_loo

Depends:

-   none

```{r}
f_name <- "target_encoding_loo"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE target_encoding_lab

Depends:

- identify_categorical_variables
- identify_response_type    
- validate_arg_df           
- validate_arg_df_not_null  
- validate_arg_encoding_method
- validate_arg_function_name 
- validate_arg_quiet

```{r}
f_name <- "target_encoding_lab"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

# PREFERENCE ORDER


## DONE case_weights

Depends:

-   validate_arg_function_name

```{r}
f_name <- "case_weights"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```


## DONE score_auc

Depends:

-   validate_arg_function_name

```{r}
f_name <- "score_auc"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE score_r2

Depends:

-   validate_arg_function_name

```{r}
f_name <- "score_r2"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE score_cramer

Depends:

-   validate_arg_function_name

```{r}
f_name <- "score_cramer"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```
## DONE f_auto_rules

Depends:

-   none

```{r}
f_name <- "f_auto_rules"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE f_functions

Depends:

-   none

```{r}
f_name <- "f_functions"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```
## DONE f_auto

Depends:

- f_auto_rules    
- identify_response_type 
- validate_arg_df        
- validate_arg_function_name
- validate_arg_predictors 
- validate_arg_quiet    
- validate_arg_responses

```{r}
f_name <- "f_auto"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
#cannot go beyond 87%
```


## DONE f_numeric_glm

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_numeric_glm"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```
## DONE f_count_glm

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_count_glm"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE f_binomial_glm

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_binomial_glm"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```



## DONE f_numeric_gam

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_numeric_gam"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE f_count_gam

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_count_gam"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE f_binomial_gam

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_binomial_gam"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```



## DONE f_numeric_rf

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_numeric_rf"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```
## DONE f_count_rf

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_count_rf"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE f_binomial_rf

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_binomial_rf"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE f_categorical_rf

Depends:

-   validate_arg_function_name

```{r}
f_name <- "f_categorical_rf"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_f

Depends:

-   validate_arg_function_name

```{r}
f_name <- "validate_arg_f"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE preference_order

Depends:

- f_functions
- validate_arg_df     
- validate_arg_df_not_null 
- validate_arg_f      
- validate_arg_function_name
- validate_arg_predictors 
- validate_arg_quiet     
- validate_arg_responses

```{r}
f_name <- "preference_order"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_preference_order

Depends:

- validate_arg_df_not_null  
- validate_arg_function_name
- validate_arg_predictors   
- validate_arg_responses 

```{r}
f_name <- "validate_arg_preference_order"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```


# CORRELATION


## DONE validate_arg_max_cor

Depends:

-   validate_arg_function_name

```{r}
f_name <- "validate_arg_max_cor"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE cor_cramer

Depends:

-   validate_arg_function_name

```{r}
f_name <- "cor_cramer"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE cor_df

Depends:

- validate_arg_df              
- validate_arg_df_not_null   
- validate_arg_function_name    
- validate_arg_predictors     
- validate_arg_quiet

```{r}
f_name <- "cor_df"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```



## DONE cor_matrix

Depends:

-   cor_df

```{r}
f_name <- "cor_matrix"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE cor_select

Depends:

- collinear_select
- validate_arg_df         
- validate_arg_df_not_null
- validate_arg_function_name  
- validate_arg_predictors 
- validate_arg_quiet 

```{r}
f_name <- "cor_select"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE cor_stats

Depends:

- cor_df
- validate_arg_df_not_null 
- validate_arg_function_name
- validate_arg_quiet 

```{r}
f_name <- "cor_stats"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE cor_clusters

Depends:

-   validate_arg_function_name
-   cor_matrix

```{r}
f_name <- "cor_clusters"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
#cannot go further
```

# VIF

## DONE vif

Depends:

-   validate_arg_function_name

```{r}
f_name <- "vif"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE validate_arg_max_vif

Depends:

-   validate_arg_function_name

```{r}
f_name <- "validate_arg_max_vif"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE vif_df

Depends:

- cor_matrix
- validate_arg_df     - validate_arg_df_not_null  
- validate_arg_function_name
- validate_arg_predictors   
- validate_arg_quiet  - vif 

```{r}
f_name <- "vif_df"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```


## DONE vif_stats

Depends:

- validate_arg_df_not_null 
- validate_arg_function_name
- vif_df

```{r}
f_name <- "vif_stats"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

```


## DONE vif_select

Depends:

- collinear_select
- validate_arg_df         
- validate_arg_df_not_null
- validate_arg_function_name  
- validate_arg_predictors 
- validate_arg_quiet 

```{r}
f_name <- "vif_select"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

#MISC

## DONE model_formula

Depends:

- identify_valid_variables
- validate_arg_df_not_null  
- validate_arg_function_name
- validate_arg_predictors   
- validate_arg_quiet        
- validate_arg_responses 

```{r}
f_name <- "model_formula"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

#COLLINEAR


## DONE collinear_stats

Depends:

- cor_df                  
- cor_matrix              
- cor_stats
- validate_arg_df         
- validate_arg_df_not_null
- validate_arg_function_name
- validate_arg_predictors   
- vif_stats

```{r}
f_name <- "collinear_stats"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE collinear_select

Depends:

- cor_matrix
- validate_arg_df              
- validate_arg_df_not_null   
- validate_arg_function_name 
- validate_arg_max_cor      
- validate_arg_max_vif      
- validate_arg_predictors    
- validate_arg_preference_order
- validate_arg_quiet        
- validate_arg_responses    
- vif

```{r}
f_name <- "collinear_select"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE class.collinear_selection

Depends:

- identify_valid_variables
- model_formula
- validate_arg_function_name
- f_functions

```{r}
f_name <- "class.collinear_selection"
f_source <- paste0("../R/S3_methods_", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE summary.collinear_selection

Depends:

```{r}
f_name <- "summary.collinear_selection"
f_source <- paste0("../R/S3_methods_", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE collinear

Depends:

- class.collinear_selection   
- collinear_select            
- cor_matrix             
- identify_valid_variables  
- target_encoding_lab          
- validate_arg_df             
- validate_arg_df_not_null    
- validate_arg_function_name  
- validate_arg_predictors   
- validate_arg_preference_order
- validate_arg_quiet          
- validate_arg_responses

```{r}
f_name <- "collinear"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```


## DONE collinear_auto

Depends:

- cor_df                  
- cor_matrix              
- cor_stats               
- data.frame             
- f_auto
- validate_arg_df         
- validate_arg_df_not_null
- validate_arg_function_name
- validate_arg_predictors  
- validate_arg_quiet    
- validate_arg_responses

```{r}
f_name <- "collinear_auto"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```

## DONE summary.collinear_output

Depends:

-   none

```{r}
f_name <- "summary.collinear_output"
f_source <- paste0("../R/S3_methods_", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

#codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE print.collinear_selection

Depends:

-   whatevs

```{r}
f_name <- "print.collinear_selection"
f_source <- paste0("../R/S3_methods_", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```

## DONE print.collinear_output

Depends:

-   whatevs

```{r}
f_name <- "print.collinear_output"
f_source <- paste0("../R/S3_methods_", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)
```


## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```

## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```



## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```


## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```


## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```


## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```


## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```


## template

Depends:

-   whatevs

```{r}
f_name <- "whatevs"
f_source <- paste0("../R/", f_name, ".R")
f_test <- paste0("../tests/testthat/test-", f_name, ".R")

codetools::findGlobals(fun = get(f_name))

testthat::test_file(f_test)

covr::file_coverage(
  source_files = f_source,
  test_files = f_test
)

covr::report(
  x = covr::file_coverage(
    source_files = f_source,
    test_files = f_test
  )
)
```
