---
title: "Adaptive Filtering Thresholds"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
    source: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Summary

The function [collinear()] configures multicollinearity filtering thresholds automatically when users don't specify them explicitly. This adaptive approach eliminates the guesswork of choosing values for the `max_cor`and `max_vif` arguments, and adapts filtering strength to the correlation structure of the data.

This article explains how the automatic threshold configuration works, and describes the experiments used to tune it.

## Setup

For this article we need the following packages and configuration. Notice the vector `predictors_high_collinearity` with a subset of `vi_predictors_numeric` with a very high collinearity.

```{r, message = FALSE, warning = FALSE, results = "hide"}
library(collinear)
library(future)
library(ggplot2)
library(patchwork)

#parallelization setup
#only useful for categorical predictors
future::plan(
  future::multisession,
  workers = future::availableCores() - 1
)

#progress bar (does not work in Rmarkdown)
#progressr::handlers(global = TRUE)

#example data
data(
  vi_smol, 
  vi_predictors_numeric,
  experiment_adaptive_thresholds
  )
```


## Adaptive Threshold Selection

The arguments `max_cor` and `max_vif` define the maximum pairwise correlation and variance inflation factors allowed during multicollinearity filtering. 

Unlike `collinear_select()`, `cor_select()`, and `vif_select()`, which use fixed defaults (`max_cor = 0.7`, `max_vif = 5`), the function `collinear()` sets these to NULL and computes appropriate thresholds from the data.

Let me show you why.

The vector `vi_predictors_numeric` names numeric predictors with a moderate multicollinearity, as the stats below show.

```{r}
collinear::collinear_stats(
  df = vi_smol,
  predictors = vi_predictors_numeric
) |> 
  dplyr::filter(
    statistic %in% c("quantile_0.75", "maximum")
  )
```

We can run them through `collinear()`.

```{r}
x <- collinear::collinear(
  df = vi_smol,
  predictors = vi_predictors_numeric
)
```

The first two messages above indicate the values of `max_cor` and `max_vif` selected by `collinear()` based on the data properties. The last message indicates the predictors selected in this run. Their stats are shown below.

```{r}
collinear::collinear_stats(
  df = x$result$df,
  predictors = x$result$selection
) |> 
  dplyr::filter(
    statistic %in% c("quantile_0.75", "maximum")
  )
```

These stats show a high median correlation but reasonable VIF scores.

Let's repeat this experiment, but pretending that `collinear()` has the default values `max_cor = 0.7` and `max_vif = 5` (same as [collinear_select()]).

```{r}
x <- collinear::collinear(
  df = vi_smol,
  predictors = vi_predictors_numeric,
  max_cor = 0.7,
  max_vif = 5
)

collinear::collinear_stats(
  df = x$result$df,
  predictors = x$result$selection
) |> 
  dplyr::filter(
    statistic %in% c("quantile_0.75", "maximum")
  )
```

Notice how this time `collinear()` returns 2 predictors (instead of 4) with a very low VIF.

Arguably, in this case `collinear()` has *over-filtered* the predictors, which might be statistically correct but is not practical for many use cases. 

By adapting thresholds to each dataset's structure, `collinear()` provides sensible defaults while still allowing manual override when needed.

## Effectiveness

To assess the effectiveness of this adaptive threshold selection method I ran `collinear()` on 10.000 random subsets of a large dataset (250 columns and 30k rows).

The results of this simulation are in the dataframe [experiment_adaptive_thresholds]. It contains the input  data size, the number of predictors selected by `collinear()`, and the collinearity stats before and after the multicollinearity filtering.

```{r, echo = FALSE, fig.width=8, fig.height=3.5}
experiment_adaptive_thresholds <- experiment_adaptive_thresholds |> 
  dplyr::mutate(
    cor_median_diff = input_cor_median - output_cor_median,
    vif_max_diff = input_vif_max - output_vif_max
  )

cor_plot <- experiment_adaptive_thresholds |>
  ggplot2::ggplot() +
  ggplot2::aes(
    x = input_cor_median,
    y = output_vif_max,
    color = (output_predictors/input_predictors)*100
  ) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::scale_color_viridis_c(option = "turbo", direction = 1) +
  ggplot2::labs(
    title = "Input Correlation vs Output VIF",
    x = "Input Median Correlation",
    y = "Output Maximum VIF",
    color = "Selected\nPredictors\n(%)"
  ) +
  ggplot2::theme_bw()

preds_plot <- experiment_adaptive_thresholds |>
  ggplot2::ggplot() +
  ggplot2::aes(
    x = input_predictors,
    y = output_predictors,
    color = output_vif_max
  ) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_smooth(method = "gam", formula = y ~ s(x, k = 3), color = "gray50", se = TRUE) +
  ggplot2::scale_color_viridis_c(option = "turbo", direction = 1) +
  ggplot2::labs(
    title = "Input vs. Selected Predictors",
    x = "Input Predictors",
    y = "Output Predictors",
    color = "Output\nmax VIF"
  ) +
  ggplot2::scale_y_continuous(breaks = c(0, 5, 10, 15, 20)) +
  ggplot2::theme_bw()

cor_plot | preds_plot
```

The simulation results shows that the adaptive threshold selection works well in most use cases, with only `r (sum(experiment_adaptive_thresholds$cor_median_diff <= 0)/nrow(experiment_adaptive_thresholds))/100`% of the instances producing suboptimal results.


## Step By Step

The adaptive selection of multicollinearity thresholds requires two steps:

1. Computation of the median correlation of the input predictors via [cor_stats()].

```{r}
max_cor <- collinear::cor_stats(
  df = vi_smol,
  predictors = predictors_high_collinearity,
  quiet = TRUE
) |> 
  dplyr::filter(statistic == "median") |> 
  dplyr::pull(value)

max_cor
```


2. The pre-trained GAM model [gam_cor_to_vif] predicts a suitable `max_vif` from the given `max_cor`. 

```{r}
max_vif <- mgcv::predict.gam(
  object = collinear::gam_cor_to_vif,
  newdata = data.frame(max_cor = max_cor)
) 

max_vif
```

Steps 1 and 2 are quite clear, but... where is the model used in step 3 coming from? 

This model was trained on data resulting from a simulation that:

 - Generated random subset of a larger numeric dataset (250 columns and 30k rows).
 - Applied [cor_select()] to the with `max_cor` value selected at random from a uniform distribution.
 - Applied [vif_select()] with increasing `max_vif` values until finding a predictor selection with the highest Jaccard similarity to the one resulting from [cor_select()].
 
The input `max_cor`, and the `max_vif` leading to a higher selection similarity were stored in the dataframe [experiment_cor_vs_vif], with `r (nrow(collinear::experiment_cor_vs_vif))` rows.

This dataframe was used to fit the model below. Notice how the `weights` argument is adjusted to the cubic transformation of the Jaccard scores to give more relevance to the cases with higher similarity between predictor selections.

```{r}
m <- mgcv::gam(
  formula = output_max_vif ~ s(input_max_cor, k = 9),
  weights = experiment_cor_vs_vif$jaccard_cor_vs_vif_selection^3,
  data = experiment_cor_vs_vif,
  select = TRUE
)

summary(m)
```

```{r}
collinear::experiment_cor_vs_vif |> 
  dplyr::arrange(
    jaccard_cor_vs_vif_selection
  ) |> 
ggplot2::ggplot() + 
  ggplot2::aes(
    x = input_max_cor,
    y = output_max_vif,
    color = jaccard_cor_vs_vif_selection,
    weight = jaccard_cor_vs_vif_selection^3
  ) + 
  ggplot2::geom_point(alpha = 0.5) + 
  ggplot2::geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 9),
    method.args = list(
      select = TRUE
    ),
    color = "black",
    se = TRUE
  ) +
  ggplot2::scale_color_viridis_c(
    option = "turbo", 
    direction = 1
    ) + 
  ggplot2::labs(
    title = "Experiment Pearson Correlation vs. VIF",
    x = "Input max_cor",
    y = "max_vif leading to most similar selection",
    color = "Jaccard\nsimilarity\nbetween\nselections"
  ) + 
  ggplot2::theme_bw()
```




```{r}
m <- mgcv::gam(
  formula = output_max_vif ~ s(input_max_cor, k = 9),
  weights = experiment_cor_vs_vif$jaccard_cor_vs_vif_selection^3,
  data = experiment_cor_vs_vif,
  select = TRUE
)

summary(m)
```


```{r, fig.width=7, fig.height=5}
# Load the equivalence data
data(equivalence_cor_vif)

# Visualize the relationship
ggplot2::ggplot(experiment_cor_vs_vif) +
  ggplot2::aes(x = max_cor, y = max_vif) +
  ggplot2::geom_point(color = "gray50") +
  ggplot2::geom_line(color = "steelblue", linewidth = 1) +
  ggplot2::geom_point(
    data = data.frame(
      max_cor = max_cor,
      max_vif = max_vif
    ),
    color = "red",
    size = 3
  ) +
  ggplot2::labs(
    x = "Maximum Correlation Threshold",
    y = "Maximum VIF Threshold",
    title = "Correlation-VIF Equivalence",
    subtitle = "Red point shows auto-configured thresholds for vi_smol dataset"
  ) +
  ggplot2::theme_bw()
```

The red point shows where our example dataset falls on this curve. Datasets with higher median correlation receive higher thresholds, while those with lower correlation get more conservative thresholds (but never below 0.58).

## Practical Examples

### Example 1: Automatic Configuration (Default)

When you don't specify thresholds, `collinear()` configures them automatically:

```{r}
result_auto <- collinear::collinear(
  df = vi_smol,
  response = "vi_numeric",
  predictors = vi_predictors,
  quiet = FALSE  # Show messages about automatic configuration
)
```

The function messages inform you of the chosen thresholds. These adapt to your specific dataset's correlation structure.

### Example 2: Manual Configuration

You can override automatic configuration by specifying one or both thresholds:

```{r}
# Specify only max_cor (max_vif is ignored, not auto-configured)
result_cor_only <- collinear::collinear(
  df = vi_smol,
  response = "vi_numeric",
  predictors = vi_predictors,
  max_cor = 0.7,
  quiet = TRUE
)

# Specify both thresholds
result_manual <- collinear::collinear(
  df = vi_smol,
  response = "vi_numeric",
  predictors = vi_predictors,
  max_cor = 0.7,
  max_vif = 5,
  quiet = TRUE
)
```

### Example 3: High vs. Low Multicollinearity Datasets

Let's compare how automatic configuration adapts to datasets with different multicollinearity levels:

```{r}
# Simulate low multicollinearity dataset
set.seed(123)
n <- 1000
df_low <- data.frame(
  y = rnorm(n),
  x1 = rnorm(n),
  x2 = rnorm(n),
  x3 = rnorm(n),
  x4 = rnorm(n)
)

# Check median correlation
cor_low <- collinear::cor_stats(
  df = df_low,
  predictors = c("x1", "x2", "x3", "x4"),
  quiet = TRUE
)

median_cor_low <- cor_low[cor_low$statistic == "median", "value"]

cat("Low multicollinearity dataset:\n")
cat("  Median correlation:", round(median_cor_low, 3), "\n")
cat("  Auto max_cor:", max(0.58, round(median_cor_low, 2)), "\n\n")

# Simulate high multicollinearity dataset
df_high <- data.frame(
  y = rnorm(n),
  x1 = rnorm(n)
)
df_high$x2 <- df_high$x1 + rnorm(n, sd = 0.3)
df_high$x3 <- df_high$x1 + rnorm(n, sd = 0.3)
df_high$x4 <- df_high$x2 + rnorm(n, sd = 0.3)

# Check median correlation
cor_high <- collinear::cor_stats(
  df = df_high,
  predictors = c("x1", "x2", "x3", "x4"),
  quiet = TRUE
)

median_cor_high <- cor_high[cor_high$statistic == "median", "value"]

cat("High multicollinearity dataset:\n")
cat("  Median correlation:", round(median_cor_high, 3), "\n")
cat("  Auto max_cor:", max(0.58, round(median_cor_high, 2)), "\n")
```

Notice how the low multicollinearity dataset gets the minimum threshold (0.58), while the high multicollinearity dataset receives a more permissive threshold adapted to its correlation structure.

## Comparison of Filtering Results

Let's examine how automatic versus manual configuration affects the final selection:

```{r}
# Get selections from different approaches
selection_auto <- result_auto$vi_numeric$selection
selection_manual <- result_manual$selection

cat("Automatic configuration:\n")
cat("  Selected predictors:", length(selection_auto), "\n")
cat("  Thresholds: max_cor =", max_cor_auto, ", max_vif =", max_vif_auto, "\n\n")

cat("Manual configuration:\n")
cat("  Selected predictors:", length(selection_manual), "\n")
cat("  Thresholds: max_cor = 0.7, max_vif = 5\n")
```

We can visualize the overlap between selections:

```{r, fig.width=7, fig.height=5}
# Create comparison data
comparison <- data.frame(
  predictor = unique(c(selection_auto, selection_manual)),
  auto = NA,
  manual = NA
)

comparison$auto <- ifelse(comparison$predictor %in% selection_auto, "Selected", "Not Selected")
comparison$manual <- ifelse(comparison$predictor %in% selection_manual, "Selected", "Not Selected")

# Count agreement
both_selected <- sum(comparison$auto == "Selected" & comparison$manual == "Selected")
only_auto <- sum(comparison$auto == "Selected" & comparison$manual == "Not Selected")
only_manual <- sum(comparison$auto == "Not Selected" & comparison$manual == "Selected")

cat("\nSelection agreement:\n")
cat("  Both methods:", both_selected, "\n")
cat("  Only automatic:", only_auto, "\n")
cat("  Only manual:", only_manual, "\n")
```

## When to Use Automatic vs. Manual Configuration

### Use Automatic Configuration When:

1. **Exploratory analysis**: You're getting familiar with a new dataset
2. **No domain knowledge**: You don't have strong priors about appropriate thresholds
3. **Multiple datasets**: You're applying the same analysis to datasets with varying correlation structures
4. **Reproducibility**: You want filtering criteria that adapt to data changes

### Use Manual Configuration When:

1. **Specific requirements**: Your field has established threshold conventions
2. **Regulatory compliance**: Your analysis must meet predetermined criteria
3. **Comparative studies**: You need consistent thresholds across multiple analyses
4. **Fine-tuning**: You want more or less aggressive filtering than automatic configuration provides

## Best Practices

1. **Start with automatic**: Begin with automatic configuration to understand your data's natural threshold
2. **Inspect the results**: Check the selected predictors and multicollinearity statistics
3. **Adjust if needed**: Use manual configuration if automatic thresholds don't meet your needs
4. **Document your choice**: Always note whether you used automatic or manual configuration in your analysis

## Performance Testing

The automatic configuration approach has been validated through extensive simulations (see `experiment_collinear_auto`):

```{r}
data(experiment_collinear_auto)

# Summary of reduction in predictors
reduction <- (experiment_collinear_auto$input_predictors - 
              experiment_collinear_auto$output_predictors) / 
             experiment_collinear_auto$input_predictors * 100

cat("Simulation results (n =", nrow(experiment_collinear_auto), "experiments):\n")
cat("  Mean predictor reduction:", round(mean(reduction), 1), "%\n")
cat("  Mean output max VIF:", round(mean(experiment_collinear_auto$output_vif_max), 2), "\n")
cat("  Mean output median correlation:", round(mean(experiment_collinear_auto$output_cor_median), 2), "\n")
```

These results demonstrate that automatic configuration consistently produces selections with low multicollinearity while retaining a substantial proportion of the original predictors.

## Recommendations

1. **Default behavior**: Trust the automatic configuration for most analyses. It's designed to work well across a wide range of datasets.

2. **Monitor messages**: Pay attention to the messages showing auto-configured thresholds. If they seem unusual for your data, investigate further.

3. **Check outputs**: Always inspect the resulting VIF scores and correlations using `vif_df()` or `cor_df()` to ensure the filtering meets your needs.

4. **Iterate if needed**: If automatic configuration is too aggressive or too permissive, switch to manual configuration with appropriate thresholds.

5. **Document your approach**: Whether using automatic or manual configuration, document your choice and reasoning for reproducibility.

## Summary

The adaptive filtering threshold feature in `collinear`:

- Eliminates guesswork by analyzing each dataset's correlation structure
- Uses a validated correlation-VIF equivalence model
- Maintains minimum filtering standards (max_cor ≥ 0.58)
- Provides flexibility through manual override options
- Has been tested across thousands of simulated scenarios

This approach makes multicollinearity management more accessible while maintaining statistical rigor, allowing researchers to focus on interpretation rather than threshold selection.

```{r cleanup, include=FALSE}
# Reset any settings if needed
```
